The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-07 rational.tex
small sentence
intro and reorganize

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) \\
  &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]

Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer using the algebraic
identity $a=a$: such a connection would be computationally modeled by the
identity function as expected. However, the algebraic proof above implements
this connection in a much more interesting way: the transformation $0 = (-a)
+ a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the money will be paid at some point, as long as the infrastructure
  can be trusted that such promises will be fulfilled. Computationally, this
  means that negative types allow us to break the dependency between the
  producer and consumer and let each compute independently.
\end{itemize}

Instead of also discussing fractional types and related work in this
introduction, we briefly summarize our main technical contributions and move
on to the heart of the paper where these topics can be explained in their
proper context. Our main contributions are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} with a notion of negative types, that
  satisfies the isomorphism $a + (-a) = 0$. The semantics of this extension
  is expressed by having a ``dual'' evaluator that reverses the flow of
  execution for negative values. 
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism $a * (1/a) = 1$. The semantics of this
  extension is expressed by introducing logical variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types.
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers.
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every programmer.
\item We relate our notions of negative and fractional types to previous work
  on continuations. We analyze and relate two apparently unrelated lines of
  work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relates continuations to
  negative types; the second~\cite{Bernardi:2010:CSL:1749618.1749689} relates
  continuations to the fractional types of the Lambek-Grishin calculus.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Advanced Programming Examples}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Field of Algebraic Numbers}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

>>>>

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) \\
  &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer using the algebraic
identity $a=a$: such a connection would be computationally modeled by the
identity function as expected. However, the algebraic proof above implements
this connection in a much more interesting way: the transformation $0 = (-a)
+ a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as the
  infrastructure can be trusted with such promises. Computationally, this
  means that negative types allow us to break the dependency between
  producers and consumers and let each compute independently.
\end{itemize}

Instead of also discussing fractional types and related work in this
introduction, we briefly summarize our main technical contributions and move
on to the heart of the paper where these topics can be explained in their
proper context. Our main contributions are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism $a + (-a) = 0$. The
  semantics of this extension is expressed by having a ``dual'' evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism $a * (1/a) = 1$. The semantics of this
  extension is expressed by introducing logical variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into one extension that includes
  both negative and fractional types. The resulting language allows any
  rational number to be used as a type. The types satisfy the same familiar
  and intuitive isomorphisms that are satisfied in the mathematical field of
  rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRev}} }
\label{sec:pi}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}
\label{sec:frac}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rational Types}
\label{sec:rat}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Advanced Programming Examples}
\label{sec:prog}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conc}

@@

Say we have not considered recursion in this paper.
